---
title: "Biking Patterns in Boston"
author: "Mette Hejberg Pedersen"
date: "4/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading packages
```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(raster)
library(sp)
```

# Loading 2019 data
```{r data}
file_name <- "../data/bluebikes_tripdata_2019.csv"
# read:
data_2019 <- read_csv(file_name, progress = FALSE)

data_2019 <- data_2019 %>% rename("start_long" = "start station longitude",
                                      "start_lat" = "start station latitude",
                                      "end_long" = "end station longitude",
                                      "end_lat" = "end station latitude",
                                      "start_station_id" = "start station id",
                                      "start_station_name" = "start station name",
                                      "end_station_name" = "end station name",
                                      "birth_year" = "birth year")
```

# Basic data explorations
Exploring basic patterns in the data that might be interesting for further exploration
```{r}
# subscribers vs. customers (single use)
data_2019 %>% filter(usertype == "Subscriber") %>% count() # returns 1988494
data_2019 %>% filter(usertype == "Customer") %>% count() # returns 534277

# genders 
data_2019 %>% filter(gender == 0) %>% count() # 277703 (female?)
data_2019 %>% filter(gender == 1) %>% count() # 1652699 (male?)
data_2019 %>% filter(gender == 2) %>% count() # 592369 (other?)

# age range
data_2019 %>% dplyr::select(birth_year) %>% range() # return 1886 2003
data_2019 %>% filter(birth_year == 1886) %>% count() # returns 3

# trip duration range
data_2019 %>% dplyr::select(tripduration) %>% range() # 61 42567137
```

There are many more subscribers which suggests that it's mostly locals and long-term visitors using the bikes. 

It seems that there's something off about the gender data, so I will not include that. 

There is also something weird about the age data - there are three entries of 1888 and just in general quite old people so again, I will not use this information either. 

There is also a massive range in trip duration but that's not necessarily an error 

On the basis of this exploration I select the columns I want to use to reduce the strain on my computer when working with such large files

# Cleaning the data prior to analysis
```{r}
cleaned_2019 <- data_2019 %>% dplyr::select(c(tripduration, starttime, stoptime, start_lat, start_long, start_station_name, end_lat, end_long, end_station_name, usertype, year, month))
```

Selecting the unique start stations in 2019 and removing outliers
```{r}
unique_start <- cleaned_2019 %>% dplyr::select(c(start_long, start_lat, start_station_name)) %>% unique()
# there was a single outlier, so I remove it
unique_start <- unique_start[!(unique_start$start_long == 0.0000|unique_start$start_lat == 0.0000), ]

# other outliers
unique_start <- unique_start %>% filter(start_station_name != "BCBS Hingham")
unique_start <- unique_start %>% filter(start_station_name != "Main St at Beacon St")
```

A quick map
```{r}
leaflet() %>% addTiles() %>% setView(lng = -71.067083, lat = 42.341145, zoom = 11.5) %>% addCircleMarkers(lng = unique_start$start_long, lat = unique_start$start_lat) %>% addScaleBar()
```

Doing the same for the end stations
```{r}
unique_end <- cleaned_2019 %>% dplyr::select(c(end_long, end_lat, end_station_name)) %>% unique()
# I do the same for the end stations
unique_end <- unique_end[!(unique_end$end_long == 0.0000|unique_end$end_lat == 0.000), ]
# another outlier
unique_end <- unique_end %>% filter(end_station_name != "BCBS Hingham")
```

# Spatial analysis 
## Figuring out where people are going 
load Boston city boundary and neighborhoods
```{r}
# neighborhoods
neighborhoods <- st_read("../data/Boston_Neighborhoods/Boston_Neighborhoods.shp")
# reproject to get it on the map
projected_neighborhoods <- st_transform(neighborhoods, 4326)
```

Plot the neighborhoods and stations together 
start stations and neighborhoods
```{r}
p_popup <- paste0("<strong>Neighborhood: </strong>", projected_neighborhoods$Name) 

p2_popup <- paste0("<strong>Start station: </strong>", unique_start$start_station_name)

leaflet(projected_neighborhoods) %>% 
  addPolygons(
    stroke = TRUE, 
    fillColor = "transparent", color = "black", weight = 3,
    popup = p_popup) %>% # add popup 
  addTiles() %>%
  setView(lng = -71.067083, lat = 42.331145, zoom = 11.5) %>% 
  addCircleMarkers(lng = unique_start$start_long,
                   lat = unique_start$start_lat,
                   popup = p2_popup) %>% 
   addScaleBar()
```

From this map, we can get an idea of where in Boston are moving, both within the larger city boundary and within neighbors of the city.

To get the intersections of the points and the neighborhoods, I create a convex hull around the points. First I transform the longitude and latitude coordinates into a sf object.
```{r}
# remove NAs if there is any
points <- unique_start %>% 
  filter(! is.na(start_long)) %>% 
  filter(! is.na(start_lat))

# transform points to sf object
points_sf <- st_as_sf(points, coords = c("start_long", "start_lat"), crs = 4326)
points_sf
# transform its crs 
points_3035 <- st_transform(points_sf, crs = 3035)
points_3035
# create convex hull
points_ch <- st_convex_hull(st_union(points_3035))
points_ch

plot(points_ch)
class(points_ch)
```

I have now created a polygon out of the start station points.

Let's get the intersection between this polygon and the neighborhoods multipolygon
```{r}
# make sure they have the same crs
neighborhoods_3035 <- st_transform(neighborhoods, crs = 3035)

# get the intersections
points_neighborhoods_intersection <- st_intersection(points_ch, st_geometry(neighborhoods_3035))

# create buffer feature of the neighborhoods
p_n_buff <- st_union(points_neighborhoods_intersection)
```

Plot the objects together
```{r}
plot(points_ch)
plot(points_neighborhoods_intersection, border='red', lwd=2, add = TRUE)
plot(p_n_buff, add = TRUE)
```

The plot shows that the neighborhoods occupy a significant amount of space within the points polygon. However, there is also a large area that the neighborhoods do not cover.

I do the same for the end stations to see if we find the same patterns.
```{r}
# remove NAs if there is any
points_end <- unique_end %>% 
  filter(! is.na(end_long)) %>% 
  filter(! is.na(end_lat))
# transform the points into an sf object 
points_end_sf <- st_as_sf(points_end, coords = c("end_long", "end_lat"), crs = 4326)
# transform their crs
points_end_3035 <- st_transform(points_end_sf, crs = 3035)
# create convex hull
points_end_ch <- st_convex_hull(st_union(points_end_3035))

# plot it
plot(points_end_ch)
class(points_end_ch)
```

The polygon looks very much identical to the start stations polyogon

Let's get the intersection
```{r}
# get the intersections
points_neighborhoods_intersection_end <- st_intersection(points_end_ch, st_geometry(neighborhoods_3035))

# create buffer feature of the neighborhoods
p_n_buff_end <- st_union(points_neighborhoods_intersection_end)
```

Plot the objects together
```{r}
plot(points_end_ch)
plot(points_neighborhoods_intersection_end, border='red', lwd=2, add = TRUE)
plot(p_n_buff_end, add = TRUE)
```

It looks very much the same for the end stations.

Once again, the map confirms what we saw in the plot. There is an area beyond the city boundary towards the north where there are a lot of points.

# Loading the 2020 data
Let's look at the 2020 data
```{r}
file_name <- "../data/bluebikes_tripdata_2020.csv"
# read:
data_2020 <- read_csv(file_name, progress = FALSE)

data_2020 <- data_2020 %>% rename("start_long" = "start station longitude",
                                      "start_lat" = "start station latitude",
                                      "end_long" = "end station longitude",
                                      "end_lat" = "end station latitude",
                                      "start_station_id" = "start station id",
                                      "start_station_name" = "start station name",
                                      "end_station_name" = "end station name",
                                      "birth_year" = "birth year")
```

# Clearning the data prior to analysis
Let's clean the 2020 data
```{r}
cleaned_2020 <- data_2020 %>% dplyr::select(c(tripduration, starttime, stoptime, start_lat, start_long, start_station_name, end_lat, end_long, end_station_name, usertype, year, month))

cleaned_2020
```

# Spatial Analysis 
## Figuring out where people are going and comparison with the 2019 data
I'll only look at the start stations
```{r}
unique_start_2020 <- cleaned_2020 %>% dplyr::select(c(start_long, start_lat, start_station_name)) %>% unique()
# there was a single outlier, so I remove it
unique_start_2020 <- unique_start_2020[!(unique_start_2020$start_long == 0.0000|unique_start_2020$start_lat == 0.0000), ]

# another outlier
unique_start_2020 <- unique_start_2020 %>% filter(start_station_name != "BCBS Hingham")

unique_end_2020 <- cleaned_2020 %>% dplyr::select(c(end_long, end_lat, end_station_name)) %>% unique()

unique_end_2020 <- unique_end_2020[!(unique_end_2020$end_long == 0.0000|unique_end_2020$end_lat == 0.0000), ]

unique_end_2020 <- unique_end_2020 %>% filter(end_station_name != "BCBS Hingham")
```

Let's plot the start stations with the neighborhoods
```{r}
# plot the 2020 start stations
p_popup <- paste0("<strong>Neighborhood: </strong>", projected_neighborhoods$Name) 

p5_popup <- paste0("<strong>Start station: </strong>", unique_start_2020$start_station_name)

leaflet(projected_neighborhoods) %>% 
  addPolygons(
    stroke = TRUE, 
    fillColor = "yellow", 
    fillOpacity = 0.8, smoothFactor = 0.5, # to make it look nicer
    popup = p_popup) %>% # add popup 
  addTiles() %>%
  setView(lng = -71.057083, lat = 42.361145, zoom = 11.5) %>% 
  addCircleMarkers(lng = unique_start_2020$start_long,
                   lat = unique_start_2020$start_lat,
                   popup = p5_popup)
```

let's look at the 2019 data for comparison
```{r}
p2_popup <- paste0("<strong>Start station: </strong>", unique_start$start_station_name)

leaflet(projected_neighborhoods) %>% 
  addPolygons(
    stroke = TRUE, 
    fillColor = "yellow", 
    fillOpacity = 0.8, smoothFactor = 0.5, # to make it look nicer
    popup = p_popup) %>% # add popup 
  addTiles() %>%
  setView(lng = -71.057083, lat = 42.361145, zoom = 11.5) %>% 
  addCircleMarkers(lng = unique_start$start_long,
                   lat = unique_start$start_lat,
                   popup = p2_popup)
```

We can see from the maps that there are areas/stations that aren't in the 2019 data which are also outside of the city boundary. These are stations towards the west of the city boundary, beyond Brighton. 

Let's look at the intersection between the points and neighborhoods in 2020
```{r}
# remove NAs if there is any
points_2020 <- unique_start_2020 %>% 
  filter(! is.na(start_long)) %>% 
  filter(! is.na(start_lat))
# transform the points into an sf object 
points_2020_sf <- st_as_sf(points_2020, coords = c("start_long", "start_lat"), crs = 4326)
# transform their crs
points_2020_3035 <- st_transform(points_2020_sf, crs = 3035)
# create convex hull
points_2020_ch <- st_convex_hull(st_union(points_2020_3035))

# plot it
plot(points_2020_ch)
class(points_end_ch)
```

```{r}
# get the intersections
points_neighborhoods_intersection_2020 <- st_intersection(points_2020_ch, st_geometry(neighborhoods_3035))

# create buffer feature of the neighborhoods
p_n_buff_2020 <- st_union(points_neighborhoods_intersection_2020)
```

Plot the objects together
```{r}
plot(points_2020_ch)
plot(points_neighborhoods_intersection_2020, border='red', lwd=2, add = TRUE)
plot(p_n_buff_2020, add = TRUE)
```

compare this the 2019 start stations
```{r}
plot(points_end_ch)
plot(points_neighborhoods_intersection_end, border='red', lwd=2, add = TRUE)
plot(p_n_buff_end, add = TRUE)
```

The two intersections look only slightly different.

Let's find the most used start station in 2019 and 2020
```{r}
# got the function from here: https://stackoverflow.com/questions/66787325/how-to-find-the-least-frequent-value-in-a-column-of-dataframe-in-r

Modemin <- function(x){
  a = table(x) # x is a column
  return(a[which.min(a)])
}
# least used start station in 2020: MTL-ECO4-01 (1 time)
Modemin(data_2020$start_station_name)
# least used start station in 2019: 8D QC Station 01 (1 time)
Modemin(data_2019$start_station_name)
# lest used end station in 2020: Mobile Temporary Station 1 (1 time)
Modemin(data_2020$end_station_name)
# lest used end station in 2019: 8D QC Station 02 (1 time)
Modemin(data_2019$end_station_name)

Modemax <- function(x){
  a = table(x) # x is a column
  return(a[which.max(a)])
}
# most used start station in 2019: MIT at Mass Ave / Amherst St (61056 times)
Modemax(data_2019$start_station_name)
# most used start station in 2020: Central Square at Mass Ave / Essex St (32668 times )
Modemax(data_2020$start_station_name)
# most used end station in 2019: MIT at Mass Ave / Amherst St (56986 times)
Modemax(data_2019$end_station_name)
# most used end station in 2020: Central Square at Mass Ave / Essex St (33493 times)
Modemax(data_2020$end_station_name)
```

So the most used stations in 2020 are used significantly less than the most used station in 2019. Interestingly, the most used start and end stations in 2019 is the same and the most used start and end station in 2020 is also the same. 

Let's find out where these stations are and plot them with leaflet
```{r}
# most used station in 2019
data_2019 %>% filter(start_station_name == "MIT at Mass Ave / Amherst St") %>% dplyr::select(c(start_long, start_lat))
leaflet(projected_neighborhoods) %>% addTiles() %>% setView(lng = -71.1, lat = 42.4, zoom = 12) %>% addCircleMarkers(lng = -71.1, lat = 42.4) %>% 
  addPolygons(stroke = TRUE, 
    fillColor = "yellow", 
    fillOpacity = 0.8, smoothFactor = 0.5, # to make it look nicer
    popup = p_popup) %>% addScaleBar()

# most used station in 2020
data_2020 %>% filter(start_station_name == "Central Square at Mass Ave / Essex St") %>% dplyr::select(c(start_long, start_lat))
```

As it turn out the two stations have the same longitude and latitude coordinates. Interestingly, the most used station is far outside the city boundary. From the station name in the 2019 data, it seems that this might be a station close to MIT, so let's find MIT's coordinates and plot it along side the point. 

```{r}
p_popup = paste0("<strong>Start station: </strong>", "MIT")
p_popup2 = paste0("<strong>institution: </strong>", "Central Square at Mass Ave / Essex St")

leaflet(projected_neighborhoods) %>% addTiles() %>% setView(lng = -71.1, lat = 42.4, zoom = 12) %>% addCircleMarkers(lng = -71.1, lat = 42.4, popup = p_popup2) %>% 
  addCircleMarkers(lng = -71.092003, lat = 42.360001, popup = p_popup, color = "red") %>% 
  addPolygons(stroke = TRUE, 
    fillColor = "yellow", 
    fillOpacity = 0.8, smoothFactor = 0.5)
```

So the central campus of MIT is not near the most used stations in 2019 and 2020, but plotting the different higher learning institutions located in Boston might yield interesting results seeing as many of the stations are clustered around this area.

Load colleges and universities
```{r}
uni <- st_read("../data/Colleges_and_Universities/Colleges_and_Universities.shp")

# reproject to get it on the map
projected_uni <- st_transform(uni, 4326)

projected_uni <- projected_uni[!(projected_uni$Longitude == 0.0000|projected_uni$Latitude == 0.0000), ]
```

Plot the colleges and universities on a map together with the start stations from 2019 and 2020
```{r}
p7_popup <- paste0("<strong>Institution: </strong>", projected_uni$Name)

leaflet() %>% addTiles() %>% 
  addCircleMarkers(lng = projected_uni$Longitude, lat = projected_uni$Latitude, color = "red", popup = p7_popup) %>% 
  addCircleMarkers(lng = unique_start$start_long, lat = unique_start$start_lat) %>% setView(lng = -71.1, lat = 42.35, zoom = 11.5) %>% addScaleBar()
```

It does in fact seem that many of the start station points are centered around the location of the colleges and universities. 

Let's make the colleges and universities into a polygon and plot it again 
```{r}
# remove NAs if there is any
points_uni <- projected_uni %>% 
  filter(! is.na(Longitude)) %>% 
  filter(! is.na(Latitude))
# transform points into sf object
points_uni_sf <- st_as_sf(points_uni, coords = c("Longitude", "Latitude"), crs = 4326)
# transform its crs 
points_uni_3035 <- st_transform(points_uni_sf, crs = 3035)
# create convex hull
points_uni_ch <- st_convex_hull(st_union(points_uni_3035))
# plot it
plot(points_uni_ch)
# give it projected crs again
points_uni_ch_3035 <- st_transform(points_uni_ch, crs = 4326)
```

Let's plot the colleges and universities polygon together the with 2019 start stations
```{r}
leaflet(points_uni_ch_3035) %>% addTiles() %>% 
  addPolygons(color = "yellow") %>% 
  addCircleMarkers(lng = unique_start$start_long, lat = unique_start$start_lat) %>% setView(lng = -71.1, lat = 42.35, zoom = 11.5)
```

So indeed, I does seem that the colleges and institutions take up a fairly large part of the space the points occupy. Let's confirm this be getting the intersection between the points-polygon and the colleges and universities-polygon
```{r}
# get the intersections
uni_points_intersection <- st_intersection(points_ch, points_uni_ch)

# create buffer feature of the neighborhoods
p_n_buff_uni <- st_union(uni_points_intersection)
```

Plot the objects together
```{r}
plot(points_ch)
plot(uni_points_intersection, add = TRUE)
plot(points_uni_ch, border = "red", add = TRUE)
```

Compare with the neighborhoods and points intersection
```{r}
plot(points_ch)
plot(points_neighborhoods_intersection, border='red', lwd=2, add = TRUE)
plot(p_n_buff, add = TRUE)
```

Indeed, the intersection of the colleges and universities occupy around the same amount of space as the neighborhoods. However, they also occupy the same space within the points-polygon, and I therefore find the intersection between the neighborhoods and the colleges and universities
```{r}
neighborhoods_uni_intersection <- st_intersection(points_uni_ch, neighborhoods_3035)

plot(points_uni_ch)
plot(neighborhoods_uni_intersection, border="red", lwd=2, add = TRUE)
plot(p_n_buff, add = TRUE)
```

So, the neighborhoods and the learning institutions occupy much of the same space. Therefore, it is not possible to reliably confirm that the the users of the bikes are primarily seeking out the learning institutions. However, the overlay of the institutions on the points, did capture a real pattern of use just like the neighborhoods did.  

Let's look at the map with the most used start/end station again
```{r}
p_popup = paste0("<strong>Start station: </strong>", "MIT")
p_popup2 = paste0("<strong>institution: </strong>", "Central Square at Mass Ave / Essex St")

# with slight changes to the view setting
leaflet(projected_neighborhoods) %>% addTiles() %>% setView(lng = -71.1, lat = 42.4, zoom = 15.5) %>% addCircleMarkers(lng = -71.1, lat = 42.4, popup = p_popup2) %>% 
  addPolygons(stroke = TRUE, 
    fillColor = "yellow", 
    fillOpacity = 0.8, smoothFactor = 0.5)
```

When zooming in, it seems that the most used station is located close to several schools, and it might therefore yield interesting results to find the location of the schools located within Boston

Here I overlay data from public and non-public schools
```{r}
public <- st_read("../data/Public_Schools/Public_Schools.shp")
public <- st_transform(public, 4326)
public <- st_as_sf(public, crs = 4326)
public <- st_convex_hull(public$geometry)
# extract coordinates
public_coordinates <- st_coordinates(public)
class(public_coordinates)
# transform to data frame
public_df <- public_coordinates %>% 
  as_tibble()


# I do the same for the non-public schools
non_public <- st_read("../data/Non_Public_Schools/Non_Public_Schools.shp")
non_public <- st_transform(non_public, 4326)
non_public <- st_as_sf(non_public, crs = 4326)
non_public <- st_convex_hull(non_public$geometry)
# extract coordinates
non_public_coordinates <- st_coordinates(non_public)
class(non_public_coordinates)
# transform to data frame
non_public_df <- non_public_coordinates %>% 
  as_tibble()

non_public_df <- non_public_coordinates %>% as_tibble()
```

Now we can plot the schools at points on the map alongside the start stations
```{r}
p_popup <- paste0("Public")
p_popup2 <- paste0("Non-public")

leaflet() %>% addTiles() %>% 
  addCircleMarkers(lng = unique_start$start_long, lat = unique_start$start_lat) %>% 
  addCircleMarkers(lng = public_df$X, lat = public_df$Y, color = "red", popup = p_popup) %>% 
  addCircleMarkers(lng = non_public_df$X, lat = non_public_df$Y, color = "yellow", popup = p_popup2) %>% setView(lng = -71.1, lat = 42.33, zoom = 11.5) %>% addScaleBar()
```

This plot seems to reflect a usage pattern dominat in the lower part of the city. There are many overlaps of the points. Let's transform the schools into a polygon and plot again. I do this by finding the collected shared space between the points
```{r}
non_public_sf <-  st_as_sf(non_public_df, coords = c("X", "Y"), crs = 4326)
class(non_public_sf)
non_public_sf

non_public_sf <- st_transform(non_public_sf, 3035)
non_public_sf

non_public_ch <- st_convex_hull(st_union(non_public_sf))
plot(non_public_ch)

plot(non_public_ch)

public_sf <- non_public_coordinates %>% 
  as_tibble() %>% 
  sf::st_as_sf(coords=c(1,2), crs = 4326)
class(public_sf)

public_sf <- st_transform(public_sf, 3035)

public_ch <- st_convex_hull(st_union(public_sf))
plot(public_ch)

schools <- st_convex_hull(st_union(non_public_ch, public_ch))
plot(schools)
```

Plot the schools with the stations points 
```{r}
plot(points_ch)
plot(schools, border = "yellow", lwd= 2, add = TRUE)
```

So the schools also take a around the same space that the neighborhoods did and the colleges and universities did 

From these analyzes, it was not possible to confirm what was located to the north of the city where a lot of the points were clustered. However, both the institutions overlay and the schools overlay revealed genuine patterns of use in different parts of the neighborhoods within Boston.