---
title: "Routes and Elevation"
author: "Mette Hejberg Pedersen"
date: "6/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(raster)
library(sp)
library(diffdf)
library(ggplot2)
```


# Loading and cleaning data
let's load the 2019 data
```{r data}
file_name <- "../data/bluebikes_tripdata_2019.csv"
# read
data_2019 <- read_csv(file_name, progress = FALSE)

data_2019 <- data_2019 %>% rename("start_long" = "start station longitude",
                                      "start_lat" = "start station latitude",
                                      "end_long" = "end station longitude",
                                      "end_lat" = "end station latitude",
                                      "start_station_id" = "start station id",
                                      "start_station_name" = "start station name",
                                      "end_station_name" = "end station name",
                                      "birth_year" = "birth year")
```

Clean the 2019 data
```{r}
# select the columns I need
cleaned_2019 <- data_2019 %>% dplyr::select(c(tripduration, starttime, stoptime, start_lat, start_long, start_station_name, end_lat, end_long, end_station_name, usertype, year, month))

cleaned_2019
```

Select the unique entries of start stations
```{r}
unique_start <- cleaned_2019 %>% dplyr::select(c(start_long, start_lat, start_station_name)) %>% unique()
# there was a single outlier, so I remove it
unique_start <- unique_start[!(unique_start$start_long == 0.0000|unique_start$start_lat == 0.0000), ]

# other outliers
unique_start <- unique_start %>% filter(start_station_name != "BCBS Hingham")
unique_start <- unique_start %>% filter(start_station_name != "Main St at Beacon St")
```

I do the same for the end stations
```{r}
unique_end <- cleaned_2019 %>% dplyr::select(c(end_long, end_lat, end_station_name)) %>% unique()
# I do the same for the end stations
unique_end <- unique_end[!(unique_end$end_long == 0.0000|unique_end$end_lat == 0.000), ]
# another outlier
unique_end <- unique_end %>% filter(end_station_name != "BCBS Hingham")
```

Let's do the same for the 2020 data
```{r}
file_name <- "../data/bluebikes_tripdata_2020.csv"
# read
data_2020 <- read_csv(file_name, progress = FALSE)

data_2020 <- data_2020 %>% rename("start_long" = "start station longitude",
                                      "start_lat" = "start station latitude",
                                      "end_long" = "end station longitude",
                                      "end_lat" = "end station latitude",
                                      "start_station_id" = "start station id",
                                      "start_station_name" = "start station name",
                                      "end_station_name" = "end station name",
                                      "birth_year" = "birth year")
```

Let's clean the 2020 data
```{r}
cleaned_2020 <- data_2020 %>% dplyr::select(c(tripduration, starttime, stoptime, start_lat, start_long, start_station_name, end_lat, end_long, end_station_name, usertype, year, month))

cleaned_2020
```

```{r}
unique_start_2020 <- cleaned_2020 %>% dplyr::select(c(start_long, start_lat, start_station_name)) %>% unique()
# there was a single outlier, so I remove it
unique_start_2020 <- unique_start_2020[!(unique_start_2020$start_long == 0.0000|unique_start_2020$start_lat == 0.0000), ]

# another outlier
unique_start_2020 <- unique_start_2020 %>% filter(start_station_name != "BCBS Hingham")

unique_end_2020 <- cleaned_2020 %>% dplyr::select(c(end_long, end_lat, end_station_name)) %>% unique()

unique_end_2020 <- unique_end_2020[!(unique_end_2020$end_long == 0.0000|unique_end_2020$end_lat == 0.0000), ]

unique_end_2020 <- unique_end_2020 %>% filter(end_station_name != "BCBS Hingham")
```

## Routes
Top routes in 2019
```{r}
# start stations 2019
unique_start %>% dplyr::select(start_station_name) %>% unique() %>%  count() # 359

# get the station names
stations_2019 <- table(data_2019$start_station_name)
# convert to tibble
stations_2019_df <- as.data.frame(stations_2019) %>% as_tibble()
# get the indices in ascending order of frequency
order(stations_2019_df$Freq)
# select the top 36
top36_2019 <- stations_2019_df[c(24, 282, 256, 207, 60, 53, 210, 244, 284, 62, 238, 205, 331, 353, 48, 218, 108, 120, 80, 89, 36, 83, 30, 252, 61, 97, 230, 228, 173, 241, 200, 22, 298, 229, 84, 227),]
# rename station name column
top36_2019 <- rename(top36_2019,
                           "start_station_name" = "Var1")
# merge data frame with longitude and latitude coordinates from data_2019
top36_2019 <- left_join(top36_2019, unique_start, by = "start_station_name")

# somehow I got 41 stations from this, but let's plot them anyway 
leaflet() %>% addTiles() %>% 
  addCircleMarkers(lng = top36_2019$start_long, lat = top36_2019$start_lat)

# end stations 2019
unique_end %>% dplyr::select(end_station_name) %>% unique() %>%  count() # 360
# get the station names
stations_2019_end <- table(data_2019$end_station_name)
# convert to tibble
stations_2019_df_end <- as.data.frame(stations_2019_end) %>% as_tibble()
# get the indices in ascending order of frequency
order(stations_2019_df_end$Freq)
# select the top 36
top36_2019_end <- stations_2019_df_end[c(207, 24, 130, 284, 331, 60,  53, 244, 209,  62, 210, 238, 205, 218,  48, 354, 108,  80,  89,  36, 252, 120,  30,  83,  61,  97, 230, 200, 228, 173, 298, 229, 241,  22,  84, 227),]
# rename station name column
top36_2019_end <- rename(top36_2019_end,
                           "end_station_name" = "Var1")
# merge data frame with longitude and latitude coordinates from data_2019
top36_2019_end <- left_join(top36_2019_end, unique_end, by = "end_station_name")

# transform to sf
coords_start_2019 <- top36_2019 %>% st_as_sf(coords = c("start_long", "start_lat"), crs = 4326)

coords_end_2019 <- top36_2019_end %>% st_as_sf(coords = c("end_long", "end_lat"), crs = 4326)

# get the coordinates
coords_2019 <- cbind(st_coordinates(coords_start_2019$geometry),st_coordinates(coords_end_2019$geometry))

# get the routes
linestrings_2019 = st_sfc(
     lapply(1:nrow(coords_2019),
           function(i){
             st_linestring(matrix(coords_2019[i,],ncol=2,byrow=TRUE))
           }))
# give them a crs
linestrings_2019 <- linestrings_2019 %>% st_set_crs(4326)
plot(linestrings_2019)

# get average length of linestring
st_length(linestrings_2019) %>% mean() # 1494.143 m = 1.5 km

# plot the routes on a map
leaflet(linestrings_2019) %>% addTiles() %>% 
  setView(lng = -71.08083, lat = 42.361145, zoom = 13) %>% 
  addPolylines() %>% 
  addCircleMarkers(lng = top36_2019_end$end_long, lat = top36_2019_end$end_lat, color = "red", popup = paste0("<strong>End station: </strong>", top36_2019_end$end_station_name)) %>% 
  addCircleMarkers(lng = top36_2019$start_long, lat = top36_2019$start_lat, popup = paste0("<strong>Start station: </strong>", top36_2019$start_station_name)) %>% addScaleBar()
```

Let's do the same for the top routes in 2020
```{r}
# start stations 2020
unique_start_2020 %>% dplyr::select(start_station_name) %>% unique() %>%  count() # 384

stations_2020_start <- table(data_2020$start_station_name)

stations_2020_df_start <- as.data.frame(stations_2020_start) %>% as_tibble()

order(stations_2020_df_start$Freq)

top38_2020_start <- stations_2020_df_start[c(22, 132, 351, 190, 270, 318, 64, 233, 266, 102, 123, 297, 214, 301, 373, 112, 261, 210, 219, 85, 89, 223, 222, 254, 258, 248, 274, 247, 182, 38, 246, 127, 65, 20, 103, 243, 96, 90),]

top38_2020_start <- top38_2020_start %>% rename("start_station_name" = "Var1")

top38_2020_start <- left_join(top38_2020_start, unique_start_2020, by = "start_station_name")

# end stations 2020
unique_end_2020 %>% dplyr::select(end_station_name) %>% unique() %>% count() # 385

stations_2020_end <- table(data_2020$end_station_name)

stations_2020_df_end <- as.data.frame(stations_2020_end) %>% as_tibble()

order(stations_2020_df_end$Freq)

top38_2020_end <- stations_2020_df_end[c(125, 132, 193, 271, 190, 233, 319, 64, 267, 123, 102, 214, 302, 298, 112, 210, 374, 262, 219, 85, 247, 89, 223, 248, 275, 255, 246, 222,  38, 259, 127, 182, 65, 20, 243, 103, 96, 90),]

top38_2020_end <- top38_2020_end %>% rename("end_station_name" = "Var1")

top38_2020_end <- left_join(top38_2020_end, unique_end_2020, by = "end_station_name")
```

```{r}
coords_start_2020 <- top38_2020_start %>% st_as_sf(coords = c("start_long", "start_lat"), crs = 4326)

coords_end_2020 <- top38_2020_end %>% st_as_sf(coords = c("end_long", "end_lat"), crs = 4326)

coords_2020 <- cbind(st_coordinates(coords_start_2020$geometry),st_coordinates(coords_end_2020$geometry))

linestrings_2020 = st_sfc(
     lapply(1:nrow(coords_2020),
           function(i){
             st_linestring(matrix(coords_2020[i,],ncol=2,byrow=TRUE))
           }))

linestrings_2020 <- linestrings_2020 %>% st_set_crs(4326)
plot(linestrings_2020)

st_length(linestrings_2020) %>% mean() # 2137.295 m = 2,1 km
```

```{r}
leaflet(linestrings_2020) %>% addTiles() %>% 
  setView(lng = -71.08083, lat = 42.361145, zoom = 13) %>% 
  addPolylines() %>% 
  addCircleMarkers(lng = top38_2020_end$end_long, lat = top38_2020_end$end_lat, color = "red", popup = paste0("<strong>End station: </strong>", top36_2019_end$end_station_name)) %>% 
  addCircleMarkers(lng = top38_2020_start$start_long, lat = top38_2020_start$start_lat, popup = paste0("<strong>Start station: </strong>", top36_2019$start_station_name)) %>% addScaleBar()
```

Seeing as I have plotted the routes in euclidian space, they do not resemble the real routes the users took very much. However, we are able to say something about where the users move between. It does not seem to be the case that there is a trend from north to south or from vest to east. Since most dots are purple, it seems that the users move within this space in all directions. However, very tentitively, the two red dots are below the river which could indicate a from north to south migration, which one could explore with more points and routes.

## Elevation

Let's explore the elevation of the city which also can have an impact of the movement opportunities the city offers.

```{r}
# read in elevation raster
elevation <- raster("../data/_ags_ff2f2033_740d_47f5_8f3e_61b68c7d95f1.tif")

# read in the neighborhoods 
neighborhoods <- st_read("../data/Boston_Neighborhoods/Boston_Neighborhoods.shp")
# reproject to get it on the map
projected_neighborhoods <- st_transform(neighborhoods, 4326)

# define colors 
pal <- colorNumeric(c("#A52A2A", "#90A926", "#55EA13"), values(elevation),
  na.color = "transparent")
# plot raster on map 
leaflet(projected_neighborhoods) %>% addTiles() %>% setView(lng = -71.067083, lat = 42.330145, zoom = 11.5) %>% addRasterImage(elevation, colors = pal, opacity = 0.8) %>% addLegend(pal = pal, values = values(elevation),
    title = "Elevation in meters above sea level") %>% 
  addPolygons(stroke = TRUE, fillColor = "transparent", color = "black", weight = 1) %>% addCircleMarkers(lng = top36_2019$start_long, lat = top36_2019$start_lat, color = "black") %>% addScaleBar()
```

There is a lot of elevation in the southern part of the city which probably also partly can explain the fact that there are less points there. The most used stations are centered in downtown Boston where there is little elevation.

Extract elevation for stations
```{r}
# remove outliers if there are any
points <- unique_start %>% 
  filter(! is.na(start_long)) %>% 
  filter(! is.na(start_lat))

# transform points to sf object
points_sf <- st_as_sf(points, coords = c("start_long", "start_lat"), crs = 4326)

# extract the elevations of the stations
points_sf$elevation  <- raster::extract(elevation, points_sf)

# Look at the points and extraction results - add polygon on top for scale
plot(points_sf["elevation"])

# transform crs of points and neighborhoods to that of elevation raster
points_elevation <- st_transform(points_sf, crs = crs(elevation, asText = TRUE))
neighborhoods_elevation <- st_transform(neighborhoods, crs = crs(elevation, asText = TRUE))

# plot the three object together
plot(elevation)
plot(points_elevation$geometry, add = TRUE)
plot(neighborhoods_elevation$geometry, border = "black", add = TRUE)

# histogram of frequency of elevation
ggplot(points_sf, aes(x = elevation)) + geom_histogram() + labs(x="Elevation", y = "Frequency") + theme_classic()

ggsave("../data_outputs/elevation_histogram.jpg")
```

The first plot of the elevation of the stations show that there are many more stations located in areas with low elevation, and the second plot with the neighborhood polygon overlay confirms this. 

We can see from the histogram that there are more points with lower elevation than higher, and from the plots that there is a trend towards the north of lower elevation and higher elevation towards the south. The histogram also tells us that there are no points in the areas with the highest elevation. The highest elevation for the points is 50 m. Therefore, Boston does have variation in elevation, however the users of the bikes do not move within the spaces of most elevation.  

Get more routes - all unique stations of 2019
```{r}
coords_start_2019_all <- unique_start %>% st_as_sf(coords = c("start_long", "start_lat"), crs = 4326)

coords_end_2019_all <- unique_end %>% st_as_sf(coords = c("end_long", "end_lat"), crs = 4326)

# find differences between the two objects
end <- coords_end_2019_all$end_station_name %>% as_tibble()
start <- coords_start_2019_all$start_station_name %>% as_tibble()

diffdf(end, start)
# remove rows with differences
coords_end_2019_all <- coords_end_2019_all[c(0-406),]
coords_end_2019_all <- coords_end_2019_all[c(0-406),]

coords_2019_all <- cbind(st_coordinates(coords_start_2019_all$geometry), st_coordinates(coords_end_2019_all$geometry))

# get all routes in 2019
linestrings_2019_all <- st_sfc(
  lapply(1:nrow(coords_2019_all),
         function(i){
           st_linestring(matrix(coords_2019_all[i,], ncol=2, byrow=TRUE))
         })
)

plot(linestrings_2019_all)

linestrings_2019_all <- linestrings_2019_all %>% st_set_crs(4326)

lengths <- st_length(linestrings_2019_all) 
lengths %>% mean() # 4811.787 == 4,8 km 
lengths %>% median() # 4411.232 == 4.4 km

#lengths <- lengths %>% as_tibble()
lengths <- lengths %>% as.numeric()
lengths <- lengths %>% as_tibble()

ggplot(lengths, aes(x = value)) + geom_histogram() + labs(x="Length in meters", y = "Frequency") + theme_classic()

ggsave("../data_outputs/routes_histogram.jpg")
```

The histogram visually confirms that the median length closely matches the mean length.
